Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-16 manuscript.tex
figure placement, whitespace fixed for fig 1.
edits to abstract, Hopscotch formatting throughout

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (Hopscotch) upstream of \emph{tb1} enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the Hopscotch element, led us to investigate the role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the Hopscotch element is segregating at a higher than expected frequency in a number of populations of both subspecies \emph{Zea mays} ssp. \emph{parviglumis} and subspecies \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the Hopscotch element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments fail to find an important role for \emph{tb1} in controlling tillering in natural populations of \emph{parviglumis}. Our findings suggest that the role of the Hopscotch in tillering in teosinte is not as straightforward as is in domesticated maize, and that other loci may play a role in observed variation in tillering in teosinte.

\jri{Check citation format for AJB. I don't think it's numbers. adjust bibliography style accordingly } \lev{I would rather send to an open access journal. I don't like that no everyone would be able to have access to paper} \jri{AJB allows open access for \$500 which is cheaper than PLoS ONE}

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis of these changes has been a focus of maize research for a number of years\citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \citep{Doebley1990, DoebleyStec1991, LukensDoebley1999}. \jri{an odd assortment of cites. why not the 1997 nature paper?} Further work showed that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citep{StuderDoebley2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte.

Tillering, or more specifically lack of tillers in teosinte, may provide an ecological advantage to plants. In species such as teosinte, where plants grow densely packed together, if plants invest more resources into growing taller and fewer resources into tillering, plants may be able to out-compete neighbors for limiting resources such as light. \jri{there must be some citations in ecology lit to support this?} In natural populations of teosinte, variation at the \emph(tb1) locus may then play a role in the ecology of teosinte.  In this study we aim to characterize the distribution of the Hopscotch insertion in \emph{parviglumis}, \emph{mexicana}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the Hopscotch element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the Hopscotch element and tillering phenotypes in a population of \emph{parviglumis}.

@@

We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{Zea mays} ssp. \emph{mexicana} populations, and 86 \emph{parviglumis} populations) and assessed the presence or absence of the Hopscotch insertion (\ref{TableS1Mays}, \ref{TableS2}). DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \cite{Kearse2012} to amplify the entire Hopscotch element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence (~5 kb amplification product) and absence (~300 bp amplification product) of the Hopscotch insertion. Two PCRs were performed for each individual, one with primers flanking the Hopscotch (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
Hop R, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, 5'-{\small CCTCCACCTCTCATGAGATCC-3'} (\ref{FigS1Locus}). Homozygotes show a single band for either the Hopscotch element (~5 kb) or the absence of the element (~300 bp), while heterozygotes are three-banded, showing a band for both the presence and absence of the Hopscotch element as well as a band for the internal primer set (\ref{FigS2Gel}). \jri{fig references still need to say "Fig" etc., all the ref function does is set the number automaticall}. When only one PCR resolved well, we scored one allele for the individual, which explains the odd number of alleles included in our analyses. We used Phusion High Fidelity Enzyme (Finnzymes, Inc.) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the Hopscotch based on band size.

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of the start of the Hopscotch and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the Hopscotch (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \citep{Wang1999, Clark2006}. \jri{is this true? the area upstream of the hopscotch? those papers just found that something upstream was the causal variant, no?} We designed the following primers using PRIMER3 \citep{RozenSkaletsky2000}: for the 5' UTR, {\small 5' GGATAATGTGCACCAGGTGT 3'} and {\small 5' GCGTGCTAGAGACACYTGTTGCT 3'}; for the 50 kb upstream region, {\small 5' TGTCCTCGCCGCAACTC 3'} and {5' TGTACGCCCGCCCCTCATCA 3'} (\ref{FigS1Locus}). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94$^{\circ}$C for 3 min, 30 cycles of 92$^{\circ}$C for 40 s, annealing for 1 min, 72$^{\circ}$C for 40 s, and a final 10 min extension at 72$^{\circ}$C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7$^{\circ}$C and 58.8$^{\circ}$C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 $\mu$ L of amplification product. This mix was placed on a thermal cycler with the following program: 37$^{\circ}$C for 30 min, 80$^{\circ}$C for 15 min, and a final cool-down step to 4$^{\circ}$C.

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE).  To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the Hopscotch insertion may be adaptive under certain environmental conditions, we looked for significant associations between the Hopscotch frequency and environmental variables using BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\citep{Coop2010}. We used genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the Hopscotch insertion coded as an additional SNP (\ref{TableS3BayEnv}). Environmental data were obtained from \url{www.worldclim.org}, the Harmonized World Soil Database and \url{www.harvestchoice.org}, and summarized by principle component analysis \citep{Pyhajarvi2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \citep{Thornton2003} to calculate pairwise F\textsubscript{ST} between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator $\hat\theta_W$; pairwise nucleotide diversity $\hat\theta_\pi$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by Hopscotch genotype we used Phylip (\url{http://evolution.genetics.washington.edu/phylip.html}) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the Hopscotch insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \citep{Chia2012}).

@@

In order to assess patterns of linkage disequilibrium (LD) around the Hopscotch element in the context of chromosomal patterns of LD we used Tassel \citep{Bradbury2007} and calculated LD between SNPs across chromosome 1 using previously published data from twelve plants each of the Ejutla A (EjuA), Ejutla B (EjuB), San Lorenzo (SLO), and La Mesa (MSA) populations \citep{Pyhajarvi2013}. We chose these populations because we had both genotyping data for the Hopscotch as well as chromosome-wide SNP data for chromosome 1. For each population we filtered the initial set of 5,897 SNPs on chromosome 1 to accept only SNPs with a minor allele frequency of at least 0.1, resulting in 1,671, 3,023, 3,122, and 2,167 SNPs for SLO, EjuB, EjuA, and MSA, respectively. We then used Tassel \citep{Bradbury2007} to calculate linkage disequilibrium ($r^{2}$) across chromosome 1 for each population.

@@

To investigate the phenotypic effects of the Hopscotch insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the Hopscotch is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the Hopscotch in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Following germination, 206 successful germinations were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer. \lev{it ended up being a combination of drip and hand watering because they dried out so much and did better when they had water on the leaves as well as in the soil}

To investigate the phenotypic effects of the Hopscotch insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the Hopscotch is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the Hopscotch in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Successful germinations (206 individuals) were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day.

Starting on day 15, we measured tillering index, the ratio of the sum of tiller lengths to the height of the plant \citep{Briggs2007}. Tillering index has been shown to be the most effective way to observe the phenotypic effects of the Hopscotch insertion on plant architecture in maize \citep{Clark2006}. Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index, or variation at \emph{tb1} (e.g. Hopscotch genotype). Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (http://www.ars.usda.gov). We genotyped individuals for the Hopscotch insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of phenotyping 1, indicating that a minimum of 71 individuals in each genotype class are needed to detect the observed effect of the Hopscotch on tillering index. \jri{do you still have these posthoc calculations?} \lev{I believe I do, otherwise they are likely in a lab meeting slide on Dropbox -- do you want them in here?} \jri{would be good to include in a github, again so we could go back and reassess how we do things in case, for example, we decide to give the greenhouse experiment a 3rd try (yes, i'm a masochist)} \lev{I would do it a 3rd time, I'm convinced something went wrong....third time with expression analyses :) }

We performed a second phenotyping experiment (phenotyping 2) in which we germinated 372 seeds of \emph{parviglumis}, choosing equally between sites previously determined to have or not have the Hopscotch insertion. Seeds were germinated and planted on day 7 post fruit-case removal into 2 gallon pots. Plants were watered twice daily, alternating between fertilized and non-fertilized water. We began phenotyping successful germinations (302) for tillering index on day 15 post fruit case removal, and phenotyped every five days until day 50. At day 50 we measured culm diameter between the third and fourth nodes. We extracted DNA and genotyped plants following the same guidelines as in phenotyping 1.

Resulting tillering index data for each genotype class did not meet the criteria for a repeated measures ANOVA, so we transformed the data using a Box-Cox transformation ($\alpha=0$) implemented in the car package in R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between Hopscotch genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. \jri{please add SAS scripts/code to a gist or something. }

@@

Genotype of the Hopscotch insertion was confirmed with two PCRs for 837 individuals. Among the 247 maize landrace accessions genotyped, all but 8 were homozygous for the presence of the insertion (\ref{TableS1Mays}, \ref{TableS2}). \jri{please fix table/figure references to say table/figure} Within our \emph{parviglumis} and \emph{mexicana} samples we found the Hopscotch insertion segregating in 37 and 4 populations \jri{this is confusing as the map shows <37 populations. we should be consistent about reporting for all pops, or all pops with n>8, etc. or at least be explicit what cutoff we are using for each result}, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico in both subspecies (\ref{Fig1Map}) \jri{the map only shows parviglumis. can we add mexicana? the text makes it seem as if both should be on the map}. We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p<0.05).  \jri{in what direction? too many hets? what's the F? } \lev{they weren't all in the same direction, do you still want me to list out?} \jri{no need to list all, but if there were sig. deviations in multiple directions in different pops, that is worth saying. maybe could list them all out in supp. table?}

\jri{please fix whitespace and black border on figure}
%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=250mm]{Fig1Map.pdf}
    \caption{Map showing the frequency of the Hopscotch allele in populations of \emph{Zea mays} ssp. \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of alleles sampled. } 
\label{Fig1Map}
  \end{center}
\end{figure*}
%--------------------------------------------

Using our Hopscotch genotyping data, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled 8 or more alleles. F\textsubscript{CT} is 0 within our dataset, and we found similar levels of F\textsubscript{ST} among populations within each subspecies (0.22) \jri{is this an average? this comes from libsequence?} and among all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi2013} (\ref{Table1Fst}).

>>>>

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the \emph{Hopscotch} element, led us to investigate the prevalence and potential role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of both subspecies \emph{Zea mays} ssp. \emph{parviglumis} and subspecies \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments fail to find an important role for \emph{tb1} in controlling tillering in natural populations of \emph{parviglumis}. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte is not as straightforward as is in domesticated maize, and that other loci may play a role in observed variation in tillering in teosinte.

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis of these changes has been a focus of maize research for a number of years\citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \citep{Doebley1990, DoebleyStec1991, LukensDoebley1999}. \jri{an odd assortment of cites. why not the 1997 nature paper?} Further work showed that the insertion of a 4.9 kb retrotransposon (\emph{Hopscotch}) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citep{StuderDoebley2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte.

Tillering, or more specifically lack of tillers in teosinte, may provide an ecological advantage to plants. In species such as teosinte, where plants grow densely packed together, if plants invest more resources into growing taller and fewer resources into tillering, plants may be able to out-compete neighbors for limiting resources such as light. \jri{there must be some citations in ecology lit to support this?} In natural populations of teosinte, variation at the \emph(tb1) locus may then play a role in the ecology of teosinte.  In this study we aim to characterize the distribution of the \emph{Hopscotch} insertion in \emph{parviglumis}, \emph{mexicana}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the \emph{Hopscotch} element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the \emph{Hopscotch} element and tillering phenotypes in a population of \emph{parviglumis}.

@@

We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{Zea mays} ssp. \emph{mexicana} populations, and 86 \emph{parviglumis} populations) and assessed the presence or absence of the \emph{Hopscotch} insertion (\ref{TableS1Mays}, \ref{TableS2}). DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \cite{Kearse2012} to amplify the entire \emph{Hopscotch} element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence (~5 kb amplification product) and absence (~300 bp amplification product) of the \emph{Hopscotch} insertion. Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
Hop R, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, 5'-{\small CCTCCACCTCTCATGAGATCC-3'} (\ref{FigS1Locus}). Homozygotes show a single band for either the \emph{Hopscotch} element (~5 kb) or the absence of the element (~300 bp), while heterozygotes are three-banded, showing a band for both the presence and absence of the \emph{Hopscotch} element as well as a band for the internal primer set (\ref{FigS2Gel}). \jri{fig references still need to say "Fig" etc., all the ref function does is set the number automaticall}. When only one PCR resolved well, we scored one allele for the individual, which explains the odd number of alleles included in our analyses. We used Phusion High Fidelity Enzyme (Finnzymes, Inc.) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the \emph{Hopscotch} based on band size.

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of the start of the \emph{Hopscotch} and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the \emph{Hopscotch} (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \citep{Wang1999, Clark2006}. \jri{is this true? the area upstream of the \emph{Hopscotch}? those papers just found that something upstream was the causal variant, no?} We designed the following primers using PRIMER3 \citep{RozenSkaletsky2000}: for the 5' UTR, {\small 5' GGATAATGTGCACCAGGTGT 3'} and {\small 5' GCGTGCTAGAGACACYTGTTGCT 3'}; for the 50 kb upstream region, {\small 5' TGTCCTCGCCGCAACTC 3'} and {5' TGTACGCCCGCCCCTCATCA 3'} (\ref{FigS1Locus}). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94$^{\circ}$C for 3 min, 30 cycles of 92$^{\circ}$C for 40 s, annealing for 1 min, 72$^{\circ}$C for 40 s, and a final 10 min extension at 72$^{\circ}$C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7$^{\circ}$C and 58.8$^{\circ}$C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 $\mu$ L of amplification product. This mix was placed on a thermal cycler with the following program: 37$^{\circ}$C for 30 min, 80$^{\circ}$C for 15 min, and a final cool-down step to 4$^{\circ}$C.

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE).  To calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the \emph{Hopscotch} insertion may be adaptive under certain environmental conditions, we looked for significant associations between the \emph{Hopscotch} frequency and environmental variables using BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\citep{Coop2010}. We used genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional SNP (\ref{TableS3BayEnv}). Environmental data were obtained from \url{www.worldclim.org}, the Harmonized World Soil Database and \url{www.harvestchoice.org}, and summarized by principle component analysis \citep{Pyhajarvi2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \citep{Thornton2003} to calculate pairwise F\textsubscript{ST} between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator $\hat\theta_W$; pairwise nucleotide diversity $\hat\theta_\pi$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by \emph{Hopscotch} genotype we used Phylip (\url{http://evolution.genetics.washington.edu/phylip.html}) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the \emph{Hopscotch} insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \citep{Chia2012}).

@@

In order to assess patterns of linkage disequilibrium (LD) around the \emph{Hopscotch} element in the context of chromosomal patterns of LD we used Tassel \citep{Bradbury2007} and calculated LD between SNPs across chromosome 1 using previously published data from twelve plants each of the Ejutla A (EjuA), Ejutla B (EjuB), San Lorenzo (SLO), and La Mesa (MSA) populations \citep{Pyhajarvi2013}. We chose these populations because we had both genotyping data for the \emph{Hopscotch} as well as chromosome-wide SNP data for chromosome 1. For each population we filtered the initial set of 5,897 SNPs on chromosome 1 to accept only SNPs with a minor allele frequency of at least 0.1, resulting in 1,671, 3,023, 3,122, and 2,167 SNPs for SLO, EjuB, EjuA, and MSA, respectively. We then used Tassel \citep{Bradbury2007} to calculate linkage disequilibrium ($r^{2}$) across chromosome 1 for each population.

@@

To investigate the phenotypic effects of the \emph{Hopscotch} insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the \emph{Hopscotch} is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the \emph{Hopscotch} in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Following germination, 206 successful germinations were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer. \lev{it ended up being a combination of drip and hand watering because they dried out so much and did better when they had water on the leaves as well as in the soil}

To investigate the phenotypic effects of the \emph{Hopscotch} insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the \emph{Hopscotch} is segregating at highest frequency (0.44) in our initial genotyping sample set. In order to maximize the likelihood of finding the \emph{Hopscotch} in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Successful germinations (206 individuals) were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day.

Starting on day 15, we measured tillering index, the ratio of the sum of tiller lengths to the height of the plant \citep{Briggs2007}. Tillering index has been shown to be the most effective way to observe the phenotypic effects of the \emph{Hopscotch} insertion on plant architecture in maize \citep{Clark2006}. Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index, or variation at \emph{tb1} (e.g. \emph{Hopscotch} genotype). Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (http://www.ars.usda.gov). We genotyped individuals for the \emph{Hopscotch} insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of phenotyping 1, indicating that a minimum of 71 individuals in each genotype class are needed to detect the observed effect of the \emph{Hopscotch} on tillering index. \jri{do you still have these posthoc calculations?} \lev{I believe I do, otherwise they are likely in a lab meeting slide on Dropbox -- do you want them in here?} \jri{would be good to include in a github, again so we could go back and reassess how we do things in case, for example, we decide to give the greenhouse experiment a 3rd try (yes, i'm a masochist)} \lev{I would do it a 3rd time, I'm convinced something went wrong....third time with expression analyses :) }

We performed a second phenotyping experiment (phenotyping 2) in which we germinated 372 seeds of \emph{parviglumis}, choosing equally between sites previously determined to have or not have the \emph{Hopscotch} insertion. Seeds were germinated and planted on day 7 post fruit-case removal into 2 gallon pots. Plants were watered twice daily, alternating between fertilized and non-fertilized water. We began phenotyping successful germinations (302) for tillering index on day 15 post fruit case removal, and phenotyped every five days until day 50. At day 50 we measured culm diameter between the third and fourth nodes. We extracted DNA and genotyped plants following the same guidelines as in phenotyping 1.

Resulting tillering index data for each genotype class did not meet the criteria for a repeated measures ANOVA, so we transformed the data using a Box-Cox transformation ($\alpha=0$) implemented in the car package in R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between \emph{Hopscotch} genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. \jri{please add SAS scripts/code to a gist or something. }

@@

Genotype of the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals. Among the 247 maize landrace accessions genotyped, all but 8 were homozygous for the presence of the insertion (\ref{TableS1Mays}, \ref{TableS2}). \jri{please fix table/figure references to say table/figure} Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 and 4 populations \jri{this is confusing as the map shows <37 populations. we should be consistent about reporting for all pops, or all pops with n>8, etc. or at least be explicit what cutoff we are using for each result}, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico in both subspecies (\ref{Fig1Map}) \jri{the map only shows parviglumis. can we add mexicana? the text makes it seem as if both should be on the map}. We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p<0.05).  \jri{in what direction? too many hets? what's the F? } \lev{they weren't all in the same direction, do you still want me to list out?} \jri{no need to list all, but if there were sig. deviations in multiple directions in different pops, that is worth saying. maybe could list them all out in supp. table?}

\jri{please fix whitespace and black border on figure}
%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[]
  \begin{center}
   \includegraphics[width=0.5\textwidth]{Fig1Map.pdf}
    \caption{Map showing the frequency of the \emph{Hopscotch} allele in populations of \emph{Zea mays} ssp. \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of alleles sampled. } 
\label{Fig1Map}
  \end{center}
\end{figure*}
%--------------------------------------------

Using our \emph{Hopscotch} genotyping data, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled 8 or more alleles. F\textsubscript{CT} is 0 within our dataset, and we found similar levels of F\textsubscript{ST} among populations within each subspecies (0.22) \jri{is this an average? this comes from libsequence?} and among all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi2013} (\ref{Table1Fst}).
